\documentclass[11pt, english, fleqn, DIV=15, headinclude, BCOR=1cm]{scrartcl}

\usepackage[
    bibatend,
    color,
]{../header}

\usepackage{booktabs}
\usepackage{pdflscape}

\usepackage{tikz}

\usepackage[tikz]{mdframed}
\newmdtheoremenv[%
    backgroundcolor=black!5,
    innertopmargin=\topskip,
    splittopskip=\topskip,
]{theorem}{Theorem}[section]

\newmdenv[%
    backgroundcolor=black!5,
    frametitlebackgroundcolor=black!10,
    roundcorner=5pt,
    skipabove=\topskip,
    innertopmargin=\topskip,
    splittopskip=\topskip,
    frametitle={Problem statement},
    frametitlerule=true,
]{problem}

\newmdenv[%
    backgroundcolor=white,
    frametitlebackgroundcolor=black!10,
    roundcorner=5pt,
    skipabove=\topskip,
    innertopmargin=\topskip,
    innerbottommargin=8cm,
    splittopskip=\topskip,
    frametitle={Side question},
    frametitlerule=true,
    %nobreak=true,
]{question}


\hypersetup{
    pdftitle=
}

\newcommand\inv{^{-1}}

\newcounter{totalpoints}
\newcommand\punkte[1]{#1\addtocounter{totalpoints}{#1}}

\newcounter{problemset}
\setcounter{problemset}{4}

\subject{physics751 -- Group Theory}
\ihead{physics751 -- Problem Set \arabic{problemset}}

\title{Problem Set \arabic{problemset}}

\publishers{Group 1 -- Patrick Matuschek}
\ofoot{Group 1 -- Patrick Matuschek}

\author{
    Martin Ueding \\ \small{\href{mailto:mu@martin-ueding.de}{mu@martin-ueding.de}}
}
\ifoot{Martin Ueding}

\ohead{\rightmark}

\begin{document}

\maketitle

\begin{question}
    During the lecture the regular representation of the symmetric group
    $\mathcal S_2$ was given. The multiplication table was written down:
    \[
        T_{\mathcal S_2} =
        \begin{pmatrix}
            e & (12) \\
            (12) & e
        \end{pmatrix}.
    \]

    From there, the representation was given as
    \[
        \tens D^{[\Gamma_\text{reg}]}(e) =
        \begin{pmatrix}
            1 & 0 \\ 0 & 1
        \end{pmatrix}
        \eqnsep
        \tens D^{[\Gamma_\text{reg}]}((12)) =
        \begin{pmatrix}
            0 & 1 \\ 1 & 0
        \end{pmatrix}.
    \]

    The definition given in the lecture was, as far as I recall,
    \[
        D_{jk}^{[\Gamma_\text{reg}]} (g_i) =
        \begin{cases}
            1 & g_j g_k = g_i \\
            0 & \text{else}
        \end{cases}.
    \]
    In words, this would mean that the representation matrices for a group
    element is a matrix that just had ones everywhere where that group element
    appears in the multiplication table, zeros everywhere else. This is fine
    with me.

    Then the lecturer said that on some occasions the multiplication table has
    to be rearranged in a way such that the identity is on the diagonal such
    that $D(e) = \mathbb 1$. The rearrangement theorem asserts that this
    rearrangement of rows and columns is possible. What concerns me is the fact
    that this rearranging might make the indexing of group elements
    ill-defined.

    The definition of multiplication table of a group $G$ with elements $g_i$
    that I remember is $T_{ij} := g_i g_j$. For this, there has so be something
    that maps the \emph{set} of group elements into an \emph{ordered tuple}
    such that $g_i$ is always the same element. Since the order itself is
    arbitrary, I do not mind when a permutation $\pi \in \mathcal S_{|G|}$ is
    applied such that
    \[
        G = \{ g_i\colon i \leq |G| \} = \{ g_{\pi(i)} \colon i \leq |G| \}
    \]
    still holds.

    We were all asked to derive the representation of $\mathcal S_3$ at home,
    so that is what I attempted.

    This is my $\mathcal S_3$ and the order in the set is the order I use for
    my indexing with $i$:
    \[
        \{ (1), (12), (13), (23), (123), (321) \}.
    \]
    So those are $g_1$ to $g_6$, in that order. Now I computed the
    multiplication table using the fast multiplication and the rearrangement
    theorem. This is the table that I got:

    \begin{tabular}{c|cccccc}
        $     $ & $(1  )$ & $(12 )$ & $(13 )$ & $(23 )$ & $(123)$ & $(321)$ \\
        \midrule
        $(1  )$ & $(1  )$ & $(12 )$ & $(13 )$ & $(23 )$ & $(123)$ & $(321)$ \\
        $(12 )$ & $(12 )$ & $(1  )$ & $(321)$ & $(123)$ & $(32 )$ & $(13 )$ \\
        $(13 )$ & $(13 )$ & $(123)$ & $(1  )$ & $(321)$ & $(12 )$ & $(23 )$ \\
        $(23 )$ & $(23 )$ & $(321)$ & $(123)$ & $(1  )$ & $(13 )$ & $(12 )$ \\
        $(123)$ & $(123)$ & $(13 )$ & $(23 )$ & $(12 )$ & $(321)$ & $(1  )$ \\
        $(321)$ & $(321)$ & $(23 )$ & $(12 )$ & $(13 )$ & $(1  )$ & $(123)$ \\
    \end{tabular}

    To get the unity elements on the diagonal only, I had to exchange the last
    two rows. One could also exchange the last two columns, but not both. This
    is the changed table.

    \begin{tabular}{c|cccccc}
        $     $ & $(1  )$ & $(12 )$ & $(13 )$ & $(23 )$ & \textbf{(123)} & \textbf{(321)} \\
        \midrule
        $(1  )$ & $(1  )$ & $(12 )$ & $(13 )$ & $(23 )$ & $(123)$ & $(321)$ \\
        $(12 )$ & $(12 )$ & $(1  )$ & $(321)$ & $(123)$ & $(32 )$ & $(13 )$ \\
        $(13 )$ & $(13 )$ & $(123)$ & $(1  )$ & $(321)$ & $(12 )$ & $(23 )$ \\
        $(23 )$ & $(23 )$ & $(321)$ & $(123)$ & $(1  )$ & $(13 )$ & $(12 )$ \\
        \textbf{(321)} & $(321)$ & $(23 )$ & $(12 )$ & $(13 )$ & $(1  )$ & $(123)$ \\
        \textbf{(123)} & $(123)$ & $(13 )$ & $(23 )$ & $(12 )$ & $(321)$ & $(1  )$ \\
    \end{tabular}

    At this point, I do not know that the meaning of $g_5$ and $g_6$ are
    supposed to be now.

    Then I tried the more formal way and took the definition. Using the
    ordering that I defined above, $D(e)$ will not be $\mathbb 1$. The only way
    I see to get it to work with the definition is to apply a permutation such
    that the definition is not
    \[
        D_{jk}^{[\Gamma_\text{reg}]} (g_i) =
        \begin{cases}
            1 & g_{\pi(j)} g_{\pi(k)} = g_{\pi(i)} \\
            0 & \text{else}
        \end{cases}
    \]
    but something like
    \[
        D_{jk}^{[\Gamma_\text{reg}]} (g_i) =
        \begin{cases}
            1 & g_{\pi(j)} g_{k} = g_{i} \\
            0 & \text{else}
        \end{cases}
    \]
    where only one element is permuted. To me, this looks like $s_j$ does not
    need to be $s_k$ even if $j = k$.

    What is the correct way of doing this?
\end{question}

\section{Pauli matrices and the quaternion group $Q_4$}

\newcommand\1{\sigma_1}
\newcommand\2{\sigma_2}
\newcommand\3{\sigma_3}

\begin{problem}
    Consider the group generated by $\iup\1$ and $\iup\2$, the so-called
    quaternion group $Q_4$.
\end{problem}

\subsection{Multiplication table}

\begin{problem}
    Write down the multiplication table of this group.
\end{problem}

The generators of this group to derive are $\iup\1$ and $\iup\2$. The
previously given multiplication law is
\[
    \sigma_i\sigma_j = \delta_{ij} + \iup \epsilon_{ijk} \sigma_k.
\]
Now with $g_1 := \iup\1$ and $g_2 := \iup\2$ we can write
\[
    g_1 g_2 = \iup^2 \1\2 = - \iup \sigma_3.
\]
I will call this element $g_3 := - \iup \sigma_3$.

\newcommand\qi{\mathbf i}
\newcommand\qj{\mathbf j}
\newcommand\qk{\mathbf k}

Since I know quaterions from \parencite[ยง11.1]{penrose-road_to_reality}, I will
now use his notation:
\[
    \qi := g_1 = \iup\1
    \eqnsep
    \qj := g_2 = \iup\2
    \eqnsep
    \qk := g_3 = -\iup\3.
\]
The identity element $e$ can be made up from $\qi^4$ or $\qj^4$ and $\qk^4$.
Therefore, the order of those elements is 4, by the way. These fulfill the
equations that Hamilton wrote onto a stone of Dublin's Brougham Bridge on
\printdate{1843-10-16} \parencite[198]{penrose-road_to_reality}:
\[
    \qi^2 = \qj^2 = \qk^2 = \qi\qj\qk = -1.
\]

The last one is not trivial to see, so I show that here:
\[
    \qi\qj\qk = \iup\1 \iup\2 [-\iup\3] = - \iup^3 \1\2\3 = - \iup^4 \1\1
    = - \1^2 = -1.
\]
So I am pretty confident that my new generators $\qi$ and $\qj$ yield a group
which is isomorphic to $Q_4$.

The multiplication table comes here:
\[
    \tens T =
    \begin{pmatrix}
         1   &  \qi &  \qj &  \qk & -1   & -\qi & -\qj & -\qk \\
         \qi & -1   &  \qk & -\qj & -\qi &  1   & -\qk &  \qj \\
         \qj & -\qk & -1   &  \qi & -\qj &  \qk &  1   & -\qi \\
         \qk &  \qj & -\qi & -1   & -\qk & -\qj &  \qi &  1   \\
        -1   & -\qi & -\qj & -\qk &  1   &  \qi &  \qj &  \qk \\
        -\qi &  1   & -\qk &  \qj &  \qi & -1   &  \qk & -\qj \\
        -\qj &  \qk &  1   & -\qi &  \qj & -\qk & -1   &  \qi \\
        -\qk & -\qj &  \qi &  1   &  \qk &  \qj & -\qi & -1  
    \end{pmatrix}.
\]
This could also be written in terms of the $\sigma_i$:
\[
    \tens T =
    \begin{pmatrix}
         e   &  \iup\1 &  \iup\2 &  -\iup\3 & -e   & -\iup\1 & -\iup\2 & \iup\3 \\
         \iup\1 & -e   &  -\iup\3 & -\iup\2 & -\iup\1 &  e   & \iup\3 &  \iup\2 \\
         \iup\2 & \iup\3 & -e   &  \iup\1 & -\iup\2 &  -\iup\3 &  e   & -\iup\1 \\
         -\iup\3 &  \iup\2 & -\iup\1 & -e   & \iup\3 & -\iup\2 &  \iup\1 &  e   \\
        -e   & -\iup\1 & -\iup\2 & \iup\3 &  e   &  \iup\1 &  \iup\2 &  -\iup\3 \\
        -\iup\1 &  e   & \iup\3 &  \iup\2 &  \iup\1 & -e   &  -\iup\3 & -\iup\2 \\
        -\iup\2 &  -\iup\3 &  e   & -\iup\1 &  \iup\2 & \iup\3 & -1   &  \iup\1 \\
        \iup\3 & -\iup\2 &  \iup\1 &  e   &  -\iup\3 &  \iup\2 & -\iup\1 & -e  
    \end{pmatrix}.
\]
Writing this out using only the generators seems like a really bad idea:
\[
    \tens T =
    \begin{pmatrix}
         [\iup\1]^4   &  \iup\1 &  \iup\2 &  \iup\1\iup\1 & [\iup\1]^2   & [\iup\1]^2\iup\1 & [\iup\1]^2\iup\2 & \iup\2\iup\1 \\
         \iup\1 & [\iup\1]^2   &  \iup\1\iup\1 & [\iup\1]^2\iup\2 & [\iup\1]^2\iup\1 &  [\iup\1]^4   & \iup\2\iup\1 &  \iup\2 \\
         \iup\2 & \iup\2\iup\1 & [\iup\1]^2   &  \iup\1 & [\iup\1]^2\iup\2 &  \iup\1\iup\1 &  [\iup\1]^4   & [\iup\1]^2\iup\1 \\
         \iup\1\iup\1 &  \iup\2 & [\iup\1]^2\iup\1 & [\iup\1]^2   & \iup\2\iup\1 & [\iup\1]^2\iup\2 &  \iup\1 &  [\iup\1]^4   \\
        [\iup\1]^2   & [\iup\1]^2\iup\1 & [\iup\1]^2\iup\2 & \iup\2\iup\1 &  [\iup\1]^4   &  \iup\1 &  \iup\2 &  \iup\1\iup\1 \\
        [\iup\1]^2\iup\1 &  [\iup\1]^4   & \iup\2\iup\1 &  \iup\2 &  \iup\1 & [\iup\1]^2   &  \iup\1\iup\1 & [\iup\1]^2\iup\2 \\
        [\iup\1]^2\iup\2 &  \iup\1\iup\1 &  [\iup\1]^4   & [\iup\1]^2\iup\1 &  \iup\2 & \iup\2\iup\1 & [\iup\1]^21   &  \iup\1 \\
        \iup\2\iup\1 & [\iup\1]^2\iup\2 &  \iup\1 &  [\iup\1]^4   &  \iup\1\iup\1 &  \iup\2 & [\iup\1]^2\iup\1 & [\iup\1]^2  
    \end{pmatrix}.
\]
I will just stick to the first one.

\begin{problem}
    Identify the order of the group $Q_4$.
\end{problem}

The multiplication table is an $8 \times 8$ matrix, so the order of this group
is 8.

\begin{problem}
    Identify the order of each of its elements.
\end{problem}

We have
\[
    1^m = 1
    \eqnsep \qi^4 = 1
    \eqnsep \qj^4 = 1
    \eqnsep \qk^4 = 1
    \eqnsep [-1]^2 = 1
    \eqnsep [-\qi]^4 = 1
    \eqnsep [-\qj]^4 = 1
    \eqnsep [-\qk]^4 = 1,
\]
where the orders can be read off the needed exponent to get the elements to 1.

\end{document}

% vim: spell spelllang=en tw=79
