\documentclass[11pt, english, fleqn, DIV=15, headinclude, BCOR=1cm]{scrartcl}

\usepackage[
    bibatend,
    color,
]{../header}

\usepackage{booktabs}
\usepackage{pdflscape}

\usepackage{tikz}

\usepackage[tikz]{mdframed}
\newmdtheoremenv[%
    backgroundcolor=black!5,
    innertopmargin=\topskip,
    splittopskip=\topskip,
]{theorem}{Theorem}[section]

\newmdenv[%
    backgroundcolor=black!5,
    frametitlebackgroundcolor=black!10,
    roundcorner=5pt,
    skipabove=\topskip,
    innertopmargin=\topskip,
    splittopskip=\topskip,
    frametitle={Problem statement},
    frametitlerule=true,
]{problem}

\newmdenv[%
    backgroundcolor=white,
    frametitlebackgroundcolor=black!10,
    roundcorner=5pt,
    skipabove=\topskip,
    innertopmargin=\topskip,
    innerbottommargin=8cm,
    splittopskip=\topskip,
    frametitle={Side question},
    frametitlerule=true,
    nobreak=true,
]{question}


\hypersetup{
    pdftitle=
}

\newcommand\inv{^{-1}}

\newcounter{totalpoints}
\newcommand\punkte[1]{#1\addtocounter{totalpoints}{#1}}

\newcounter{problemset}
\setcounter{problemset}{3}

\subject{physics751 -- Group Theory}
\ihead{physics751 -- Problem Set \arabic{problemset}}

\title{Problem Set \arabic{problemset}}

\publishers{Group 1 -- Patrick Matuschek}
\ofoot{Group 1 -- Patrick Matuschek}

\author{
    Martin Ueding \\ \small{\href{mailto:mu@martin-ueding.de}{mu@martin-ueding.de}}
}
\ifoot{Martin Ueding}

\ohead{\rightmark}

\begin{document}

\maketitle

Since we are likely to not cover the homework during the exercises, I added
some questions, some rather general. I would really appreciate if you could
write a bit for each.

\section{Pauli matrices}

\newcommand\1{\sigma_1}
\newcommand\2{\sigma_2}
\newcommand\3{\sigma_3}
\newcommand\ssep{,\;}

\subsection{Formation of a group}

\begin{problem}
    Show that all possible products generated by $\1$ and $\2$ form a group.
\end{problem}

The identity element of this group is the $\mathbb 1_2$ identity matrix.

To obtain the group structure, I need to find out how the products of the
elements behave.

The first things to notice is
\[
    \1\1 = e
    \eqnsep
    \2\2 = e.
\]
Then
\[
    \1\2 = \iup \3
    \eqnsep
    \2\1 = - \iup \3.
\]
Those can be used to yield
\[
    \1\2\1 = \iup \3 \1 = - \2
    \eqnsep
    \2\1\2 = - \iup \3 \2 = \iup \2\3 = - \1.
\]
The last element needed is
\[
    \1\2\1\2 = - \3\3 = - e
    \eqnsep
    \2\1\2\1 = - \3\3 = - e.
\]

There is only so much staggering with the group elements. (Infinite) products
of $\1$ and $\2$ will come down to at most 7 factors. When two equal factors
are next to each other, they reduce to the identity, and $[\1\2]^4$ is the
identity as well. This means closure.

Then there are inverses for any group element because $\1$ and $\2$ are their
own inverses. The inverse of a product is just the reversed order of elements.
Whatever way the element and its inverse is combined, there will be squares of
$\1$ and $\2$ which are the identity, until the whole expression reduced to
unity.

The other way is to work with the representation in form of the matrices.
Matrices have a multiplication law that is non-commutative and associative. The
group is supposed to be made up of all possible products of $\1$ and $\2$, so
by that definition closure is baked in.

Another way to see that it forms a group is that there exists a multiplication
table for it (see below), which implies closure and the existence of an inverse
since there is an identity in each row and column.

\subsection{Multiplication table and orders}

\begin{problem}
    Write down the multiplication table of this group and identify its order,
    the order of each element, the classes, the invariant subgroups and their
    quotient groups.
\end{problem}

\subsubsection{Multiplication table}

There are two ways now: Building up everything out of the generators $\1$ and
$\2$ as well as the identity $e$ or introducing new names for some of the
products.

The variant in table~\ref{tab:generators} uses only the generators. And in
table~\ref{tab:symbols} I have introduced $-e := \1\2\1\2$.

\begin{table}[htbp]
    \centering
    \begin{tabular}{c|ccccccc}
        $e$ & $\1$ & $\2$ & $\1\2$ & $\2\1$ & $-\1$ & $-\2$ & $-e$ \\
        \midrule
        $\1$ & $e$ & $\1\2$ & $\2$ & $-\2$ & $-e$ & $-e$ & $-\1$ \\
        $\2$ & $\2\1$ & $e$ & $-\1$ & $\1$ & $\1\2$ & $-e$ & $-\2$ \\
        $\1\2$ & $-\2$ & $\1$ & $-e$ & $e$ & $\2$ & $-\1$ & $\2\1$ \\
        $\2\1$ & $\2$ & $-\1$ & $e$ & $-e$ & $-\2$ & $\1$ & $\1\2$ \\
        $-\1$ & $-e$ & $\2\1$ & $-\2$ & $\2$ & $e$ & $\1\2$ & $\1$ \\
        $-\2$ & $\1\2$ & $-e$ & $\1$ & $-\1$ & $\2\1$ & $e$ & $\2$ \\
        $-e$ & $-\1$ & $-\2$ & $\2\1$ & $\1\2$ & $\1$ & $\2$ & $e$ \\
    \end{tabular}
    \caption{%
        Multiplication table with $-e := \1\2\1\2$ introduced.
    }
    \label{tab:symbols}
\end{table}

\begin{landscape}
    \begin{table}[p]
        \centering
        \begin{tabular}{c|ccccccc}
            $e$ & $\1$ & $\2$ & $\1\2$ & $\2\1$ & $\2\1\2$ & $\1\2\1$ & $\1\2\1\2$ \\
            \midrule
            $\1$ & $e$ & $\1\2$ & $\2$ & $\1\2\1$ & $\1\2\1\2$ & $\1\2\1\2$ & $\2\1\2$ \\
            $\2$ & $\2\1$ & $e$ & $\2\1\2$ & $\1$ & $\1\2$ & $\1\2\1\2$ & $\1\2\1$ \\
            $\1\2$ & $\1\2\1$ & $\1$ & $\1\2\1\2$ & $e$ & $\2$ & $\2\1\2$ & $\2\1$ \\
            $\2\1$ & $\2$ & $\2\1\2$ & $e$ & $\1\2\1\2$ & $\1\2\1$ & $\1$ & $\1\2$ \\
            $\2\1\2$ & $\1\2\1\2$ & $\2\1$ & $\1\2\1$ & $\2$ & $e$ & $\1\2$ & $\1$ \\
            $\1\2\1$ & $\1\2$ & $\1\2\1\2$ & $\1$ & $\2\1\2$ & $\2\1$ & $e$ & $\2$ \\
            $\1\2\1\2$ & $\2\1\2$ & $\1\2\1$ & $\2\1$ & $\1\2$ & $\1$ & $\2$ & $e$ \\
        \end{tabular}
        \caption{%
            Multiplication table built up from the generators only.
        }
        \label{tab:generators}
    \end{table}
\end{landscape}

\subsubsection{Order of the group}

There are 8 elements in the whole thing:
\[
    \{ e\ssep \1\ssep \2\ssep \1\2\ssep \2\1\ssep \2\1\2\ssep \1\2\1\ssep \1\2\1\2 \}.
\]
Therefore, the order is 8.

\subsubsection{Orders of the elements}

These are the powers needed to get each element back to unity:
\begin{gather*}
    \1^2 = e \\
    \2^2 = e \\
    [\1\2]^4 = [\iup \3]^4 = [-e]^2 = e \\
    [\2\1]^4 = [-\iup \3]^4 = [-e]^2 = e \\
    [\1\2\1]^2 = \1\2\1 \1\2\1 = e \\
    [\2\1\2]^2 = \2\1\2 \2\1\2 = e \\
    [\1\2\1\2]^2 = [\1\2\1\2]^4 = e
\end{gather*}

The orders are written down in table~\ref{tab:orders}.

\begin{table}[htbp]
    \centering
    \begin{tabular}{cc}
        Element & Order \\
        \midrule
        $\1$ & 2 \\
        $\2$ & 2 \\
        $\1\2$ & 4 \\
        $\2\1$ & 4 \\
        $\1\2\1$ & 2 \\
        $\2\1\2$ & 2 \\
        $\1\2\1\2$ & 2
    \end{tabular}
    \caption{%
        Orders of the elements.
    }
    \label{tab:orders}
\end{table}

\subsubsection{Classes}

I assume “classes” means “conjugacy classes”?

The definition of conjugacy of $a, b \in G$ is
\[
    a \sim b
    \iff
    \exists g \in G \colon
    a = g b g\inv.
\]

What I did is to go through the elements $b \in G$ and iterate through the $g
\in G$ and see which unique $a$ I could generate with that. These are the
results:
\begin{gather*}
    [e] = \{ e \} \\
    [\1] = \{ \1 \ssep \2\1\2 \} \\
    [\2] = \{ \2 \ssep \1\2\1 \} \\
    [\1\2] = \{ \1\2 \ssep \2\1 \} \\
    [\1\2\1\2] = \{ \1\2\2\1 \}
\end{gather*}

\begin{question}
    Is that way of computing the conjugacy classes recommended or is there a
    more efficient one?
\end{question}

\subsubsection{Invariant subgroup}

Invariant subgroups are made from conjugacy classes. $[e]$ has to be in it in
any case. I tried to start with $[\1]$ or $[\2]$, but the groups were either
not normal, no groups at all or the whole group itself. So I started with the
other two and got a normal subgroup:
\[
    H := \{ e \ssep \1\2 \ssep \2\1 \ssep \1\2\1\2 \}.
\]

Then I tried to pre and post multiply it with $\1$ and got the same cosets. By
analogy, this works with $\2$ as well. Doing the same with $\1\2$ and $\2\1$ is
trivial since they are contained in the group, right?

\begin{question}
    Is there some algorithm to construct invariant subgroups from conjugacy
    classes? One which does not require checking of the union actually is an
    invariant subgroup.
\end{question}

\subsubsection{Quotient group}

The definition of the quotient group is:
\[
    G / H := \{ gH \colon g \in G \}.
\]

In this case, we have the following left cosets:
\begin{gather*}
    \{ \1\ssep \1\1\2\ssep \1\2\1\ssep \1\1\2\1\2 \} = \{ \1\ssep \2\ssep \1\2\1\ssep \2\1\2 \} \\
    \{ \2\ssep \2\1\ssep \2\2\1\ssep \2\1\2\1\2 \} = \{ \1\ssep \2\ssep \1\2\1\ssep \2\1\2 \} \\
    \{ \1\2\ssep \1\2\1\ssep \1\2\2\1\ssep \1\2\1\2\1\2 \} = \{ e\ssep \1\2\ssep \2\1\ssep \1\2\1\2 \}
    \\
    \{ \2\1\ssep \2\1\1\2\ssep \2\1\2\1\ssep \2\1\1\2\1\2 \} = \{ e\ssep \1\2\ssep \2\1\ssep \1\2\1\2 \}
    \\
    \{ \1\2\1\ssep \1\2\1\1\1\2\ssep \1\2\1\2\1\ssep \1\2\1\1\2\1\2 \} \\ \qquad = \{ \1\ssep \2\ssep \1\2\1\ssep
    \2\1\2 \} \\
    \{ \2\1\2\ssep \2\1\2\1\2\ssep \2\1\2\2\1\ssep \2\1\2\1\2\1\2 \} = \{ \1\ssep \2\ssep \1\2\1\ssep
    \2\1\2 \} \\
    \{ \1\2\1\2\ssep \1\2\1\2\1\2\ssep \1\2\1\2\2\1\ssep \1\2\1\2\1\2\1\2 \} \\ \qquad = \{ e\ssep \1\2\ssep
    \2\1\ssep \1\2\1\2 \}
\end{gather*}

There are only two unique sets:
\[
    \{ e\ssep \1\2\ssep \2\1\ssep \1\2\1\2 \}
    \eqnsep
    \{ \1\ssep \2\ssep \1\2\1\ssep \2\1\2 \},
\]
which matches with my understanding of the Laplace's theorem.

\begin{question}
    Is there a shorter way than generating all the cosets and only take the
    unique ones? It seems wasteful to me to generate all and discard most.
\end{question}

\begin{question}
    In the lecture we had a theorem which said that the cosets of a normal
    subgroup are disjoint. Then it said that a group can be written as a union
    of these disjoint cosets. Are the cosets in the quotient groups those
    disjoint cosets that make up the group if united? Is that the origin of the
    name “factor group”?
\end{question}

\subsection{Isomorphy}

\begin{problem}
    Prove that this group is isomorphic to $D_4$.
\end{problem}

Given the rotation $c$ and the reflection $b$ the elements of $D_4$ are:
\[
    D_4 = \{ e\ssep c\ssep c^2\ssep c^3\ssep b\ssep bc\ssep bc^2\ssep bc^3 \}.
\]
Those are 8 elements in total, meaning that the isomorphy is not ruled out
yet.

There is a normal subgroup in $D_4$:
\[
    \{ e\ssep c\ssep c^2\ssep c^3 \}.
\]
The quotient groups is
\[
    \{ e\ssep c\ssep c^2\ssep c^3 \}
    \eqnsep
    \{ b\ssep bc\ssep bc^2\ssep bc^3 \}.
\]

The coset with the identity has to match the other one for isomorphy. So
\[
    \{ e\ssep c\ssep c^2\ssep c^3 \}
    \triangleq
    \{ e\ssep \1\2\ssep \2\1\ssep \1\2\1\2 \}
\]
With the following rearrangement, one can see that $c \triangleq \1\2$ and
therefore the following order is better:
\[
    \{ e\ssep c\ssep c^2\ssep c^3 \}
    \triangleq
    \{ e\ssep \1\2\ssep \1\2\1\2\ssep \2\1 \}.
\]

The other cosets map like this:
\[
    \{ b\ssep bc\ssep bc^2\ssep bc^3 \}
    \triangleq
    \{ \2\ssep \2\1\2\ssep \1\2\1\ssep \1 \}.
\]

Now I can see the isomorphism between all the elements. The elements match up
column wise:
\[
    \begin{matrix}
        e & \1 & \2 & \1\2 & \2\1 & \1\2\1 & \2\1\2 & \1\2\1\2 \\
        e & bc^3 & b & c & c^3 & bc^2 & bc & c^2 \\
    \end{matrix}
\]
There are other ways to chose this isomorphism though, since one could redefine
$b := bc^k$ for any $k$ and still get the same result. The existence of one
isomorphism means that they are isomorphic, though.

\end{document}

% vim: spell spelllang=en tw=79
