\documentclass[11pt, english, fleqn, DIV=15, headinclude, BCOR=1cm]{scrartcl}

\usepackage[
    bibatend,
    color,
]{../header}

\usepackage{booktabs}
\usepackage{pdflscape}

\usepackage{tikz}

\usepackage[tikz]{mdframed}
\newmdtheoremenv[%
    backgroundcolor=white,
    innertopmargin=\topskip,
    splittopskip=\topskip,
]{theorem}{Theorem}[section]

\newmdenv[%
    backgroundcolor=SframeBackground,
    frametitlebackgroundcolor=SframeTitleBackground,
    roundcorner=5pt,
    skipabove=\topskip,
    innertopmargin=\topskip,
    splittopskip=\topskip,
    frametitle={Problem statement},
    frametitlerule=true,
]{problem}

\newmdenv[%
    backgroundcolor=white,
    frametitlebackgroundcolor=SframeTitleBackground,
    roundcorner=5pt,
    skipabove=\topskip,
    innertopmargin=\topskip,
    innerbottommargin=8cm,
    splittopskip=\topskip,
    frametitle={Side question},
    frametitlerule=true,
    %nobreak=true,
]{question}

\DeclareMathOperator{\Tr}{Tr}


\hypersetup{
    pdftitle=
}

\newcommand\inv{^{-1}}

\newcounter{totalpoints}
\newcommand\punkte[1]{#1\addtocounter{totalpoints}{#1}}

\newcounter{problemset}
\setcounter{problemset}{8}

\subject{physics751 -- Group Theory}
\ihead{physics751 -- Problem Set \arabic{problemset}}

\title{Problem Set \arabic{problemset}}

\publishers{Group 1 -- Patrick Matuschek}
\ofoot{Group 1 -- Patrick Matuschek}

\author{
    Martin Ueding \\ \small{\href{mailto:mu@martin-ueding.de}{mu@martin-ueding.de}}
}
\ifoot{Martin Ueding}

\ohead{\rightmark}

\begin{document}

\maketitle

\section{The character table of $S_4$}

\subsection{Parity}

\begin{problem}
    Identify the two 1-dim.\ representations associated with the parity of the
    elements of $S_4$.
\end{problem}

The first one dimensional representation ($\Gamma_1$) is the trivial one.
Everything gets mapped to 1. The second one ($\Gamma_2$) could be the
representation that maps to 1 and $-1$ depending on the parity. This means that
\begin{align*}
    D^{\Gamma_2}(K_1)& = 1 \\
    D^{\Gamma_2}(K_2)& = -1 \\
    D^{\Gamma_2}(K_3)& = 1 \\
    D^{\Gamma_2}(K_4)& = 1 \\
    D^{\Gamma_2}(K_5)& = -1
\end{align*}

\subsection{Real characters}

\begin{problem}
    Show that the characters of all irreducible unitary representations of
    $S_n$ are real.
\end{problem}

\begin{proof}
    The inverse of any permutation in $S_n$ has the same cycle structure as the
    permutation itself. All elements of the same cycle structure are in the
    same conjugacy class. Each conjugacy class has one character, which is the
    same for every contained element. Therefore, $g$ and $g\inv$ have the same
    character. This means
    \begin{align*}
        \Tr\del{\tens D(g)} &= \Tr\del{\tens D(g\inv)}. \\
        \intertext{%
            The representation is taken to be unitary. This means that we can
            write the representation matrix of the inverse as the hermitian
            adjoint.
        }
        \Tr\del{\tens D(g)} &= \Tr\del{\tens D(g)^\dagger} \\
        \intertext{%
            The trace of the adjoint is just complex conjugated.
        }
        \Tr\del{\tens D(g)} &= \Tr\del{\tens D(g)}^* \\
        \intertext{%
            Both sides feature the same trace, the character of $g$.
        }
        \chi(g) &= \chi(g)^*
    \end{align*}
    This means that the characters are real.
\end{proof}

\subsection{Dimensions}

There are five conjugacy classes, therefore there are five irreducible
representations. The order of the group is 24. The only partition I could come
up with is
\[
    24 = 1 + 1 + 2^2 + 3^3 + 3^3.
\]

\subsection{Character of $K_3$}

The products are
\begin{gather*}
    (12)(34)(12)(34) = e \\
    (12)(34)(13)(24) = (14)(23) \\
    (12)(34)(14)(23) = (13)(24) \\
    (13)(24)(12)(34) = (14)(23) \\
    (13)(24)(13)(24) = e \\
    (13)(24)(14)(23) = (12)(34) \\
    (14)(23)(12)(34) = (13)(24) \\
    (14)(23)(13)(24) = (12)(34) \\ 
    (14)(23)(14)(23) = e.
\end{gather*}
In total, the product is
\[
    K_3 K_3 = 3 K_1 + 2 K_3.
\]
Therefore
\[
    c_{33}^1 = 3
    \eqnsep
    c_{33}^3 = 2.
\]

\begin{problem}
    Hint: Use
    \[
        |K_i| |K_j| \chi^{(\nu)}(K_i) \chi^{(\nu)}(K_j)
        = d_\nu \sum_k c_{ij}^k D^{(\nu)}(K_k)
    \]
    from your lecture.
\end{problem}

The formula that is given in the hint seems to have scalars on the left side
and the matrices (the representation $D^{(\nu)}(K_k)$) on the right. To me,
having the character on the right side as well makes more sense.

I found \parencite[(4.64)]{Luedeling/physics751-notes}, which is
\[
    n_a n_b \chi_{(\mu)}^a \chi_{(\mu)}^b = \chi_{(\mu)}^1 \sum_c c_{abc} n_c
    \chi_{(\mu)}^c.
\]
So $D^{(\nu)}(K_k)$ needs to be replaced by $|K_k| \chi^{(\nu)}(K_k)$ in the
notation from the exercise sheet.

Setting $i = j
= 3$ into the equation, this gives me
\[
    \chi^{\Gamma_3}_3 = \frac{2 \pm 4}{3}
    \eqnsep
    \chi^{\Gamma_3}_4 = 1 \pm 2
    \eqnsep
    \chi^{\Gamma_3}_5 = 1 \pm 2.
\]
Using the normalization for this column, only $2$, $-1$, $-1$ are viable
choices.

The following table summarizes everything I have learned about this group so
far:

\begin{tabular}{c|ccccc}
    & $\chi_1$ & $\chi_2$ & $\chi_3$ & $\chi_4$ & $\chi_5$ \\
    \midrule
    $|K_k|$ & 1 & 6 & 3 & 8 & 6 \\
    \midrule
    $\Gamma_1$ & 1 & 1 & 1 & 1 & 1 \\
    $\Gamma_2$ & 1 & $-1$ & 1 & 1 & $-1$ \\
    $\Gamma_3$ & 2 & & $2$ & & \\
    $\Gamma_4$ & 3 & & $-1$ & & \\
    $\Gamma_5$ & 3 & & $-1$ & & \\
\end{tabular}

\subsection{Complete character table}

The product of $K_3$ and $K_4$ is a bit lengthy:
\begin{gather*}
    (12)(34) (123) = (243) \\
    (12)(34) (132) = (143) \\
    (12)(34) (124) = (234) \\
    (12)(34) (142) = (134) \\
    (12)(34) (134) = (142) \\
    (12)(34) (143) = (132) \\
    (12)(34) (234) = (124) \\
    (12)(34) (243) = (123) \\
    %
    (13)(24) (123) = (142) \\
    (13)(24) (132) = (234) \\
    (13)(24) (124) = (143) \\
    (13)(24) (142) = (123) \\
    (13)(24) (134) = (243) \\
    (13)(24) (143) = (124) \\
    (13)(24) (234) = (132) \\
    (13)(24) (243) = (134) \\
    %
    (14)(23) (123) = (134) \\
    (14)(23) (132) = (124) \\
    (14)(23) (124) = (132) \\
    (14)(23) (142) = (243) \\
    (14)(23) (134) = (123) \\
    (14)(23) (143) = (234) \\
    (14)(23) (234) = (143) \\
    (14)(23) (243) = (142)
\end{gather*}
When all elements are taken together, this is just
\[
    K_3 K_4 = 3 K_4.
\]
This means that the class constants are just $c_{34}^4 = 3$, all other
$c_{34}^k = 0$. With
\begin{align*}
    |K_3| |K_4| \chi^{\Gamma_\nu}_3 \chi^{\Gamma_\nu}_4 &= d_\nu c_{34}^3
    |K_4| \chi^{\Gamma_\nu}_4 \\
    \iff 3 \cdot 8 \chi^{\Gamma_\nu}_3 \chi^{\Gamma_\nu}_4 &= d_\nu 3 \cdot 8
    \chi^{\Gamma_\nu}_4 \\
    \iff \sbr{\chi^{\Gamma_\nu}_3 - d_\nu} \chi^{\Gamma_\nu}_4 &= 0 \\
\end{align*}
In all the cases, where the $\chi_3$ is not the same as $\chi_1$ the character
of the fourth class is zero. The one remaining element can be deduced from
normalization to be $\pm 1$.

\begin{tabular}{c|ccccc}
    & $\chi_1$ & $\chi_2$ & $\chi_3$ & $\chi_4$ & $\chi_5$ \\
    \midrule
    $|K_k|$ & 1 & 6 & 3 & 8 & 6 \\
    \midrule
    $\Gamma_1$ & 1 & 1 & 1 & 1 & 1 \\
    $\Gamma_2$ & 1 & $-1$ & 1 & 1 & $-1$ \\
    $\Gamma_3$ & 2 & & $2$ & $\pm 1$ & \\
    $\Gamma_4$ & 3 & & $-1$ & 0 & \\
    $\Gamma_5$ & 3 & & $-1$ & 0 & \\
\end{tabular}

Looking at the normalization of the columns and rows, I can make the following
statements:
\begin{itemize}
    \item
        In the second column, the squared sum must be $24/6=4$. This mean
        that either one $\pm 2$ or two $\pm 1$ must fill the remaining places.
    \item
        Same for the fifth column. Column two and five have to be orthogonal,
        though.
    \item
        The sum in the third row as needed for the normalization,
        taking only the known places into account, is $1\cdot2^2 + 3\cdot2^2 +
        8\cdot1^2 = 4 + 12 + 8 = 24$. This already is $|G| = 24$, such that
        only zeros can be in this row.
    \item
        The sum in the fourth row is $1\cdot3^2 + 3\cdot1^2 = 12$. There is 12
        missing to make it to 24. Looking at the weighting factors (the number
        of elements in the conjugacy class), both entries have to be $\pm1$.
    \item
        The same argument applies to the fifth row, there can only be $\pm 1$
        in the gaps as well.
\end{itemize}
Using those statements, I can fill up the table like this:

\begin{tabular}{c|ccccc}
    & $\chi_1$ & $\chi_2$ & $\chi_3$ & $\chi_4$ & $\chi_5$ \\
    \midrule
    $|K_k|$ & 1 & 6 & 3 & 8 & 6 \\
    \midrule
    $\Gamma_1$ & 1 & 1 & 1 & 1 & 1 \\
    $\Gamma_2$ & 1 & $-1$ & 1 & 1 & $-1$ \\
    $\Gamma_3$ & 2 & 0 & $2$ & $\pm 1$ & 0 \\
    $\Gamma_4$ & 3 & $\pm 1$ & $-1$ & 0 & $\pm 1$ \\
    $\Gamma_5$ & 3 & $\pm 1$ & $-1$ & 0 & $\pm 1$ \\
\end{tabular}

Now those signs have to be sorted out. This can be done using the orthogonality
of the columns and rows with each other. Row four and five, as well as column
two and five are equal and have the same weighting factor. Therefore, there is
a freedom of choice here, where to have the minus signs. The sign in the third
row and fourth column has to be a minus in order to be orthogonal to the other
entries.

This is the full character table:

\begin{tabular}{c|ccccc}
    & $\chi_1$ & $\chi_2$ & $\chi_3$ & $\chi_4$ & $\chi_5$ \\
    \midrule
    $|K_k|$ & 1 & 6 & 3 & 8 & 6 \\
    \midrule
    $\Gamma_1$ & 1 & 1 & 1 & 1 & 1 \\
    $\Gamma_2$ & 1 & $-1$ & 1 & 1 & $-1$ \\
    $\Gamma_3$ & 2 & 0 & $2$ & $- 1$ & 0 \\
    $\Gamma_4$ & 3 & $1$ & $-1$ & 0 & $-1$ \\
    $\Gamma_5$ & 3 & $-1$ & $-1$ & 0 & $1$ \\
\end{tabular}

\end{document}

% vim: spell spelllang=en tw=79
